---
title: "Predict activity quality from activity monitors"
author: "Riddhi"
date: "23 October 2021"
output:
  html_document:
    keep_md: yes
---

##Synopsis

It is now possible to collect a great quantity of data regarding personal activity for a low cost using devices like the Jawbone Up, Nike FuelBand, and Fitbit. These devices are part of the quantified self movement, which is a collection of people who measure themselves on a daily basis in order to enhance their health, uncover patterns in their behaviour, or simply because they are tech geeks. People frequently quantify how much of a certain activity they engage in, but they rarely define how well they engage in it. The purpose of this experiment is to use accelerometer data from 6 participants' belts, forearms, arms, and dumbells. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways.

The purpose of this project is to forecast how they completed the activity. In the training set, this is the 'classes' variable.
## Data description

The result variable is 'classes,' a 5-level factor variable. Participants were asked to do one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in five different ways for this data set:

- exactly according to the specification (Class A)
- throwing the elbows to the front (Class B)
- lifting the dumbbell only halfway (Class C)
- lowering the dumbbell only halfway (Class D)
- throwing the hips to the front (Class E)

## Initial configuration

The initial configuration consists of loading some required packages and initializing some variables.

```{r configuration, echo=TRUE, results='hide'}

#Data variables
training.file   <- '/Users/riddhi/Downloads/pml-training.csv'
test.cases.file <- '/Users/riddhi/Downloads/pml-testing.csv'
training.url    <- 'http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv'
test.cases.url  <- 'http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv'

#Directories
if (!file.exists("data")){
  dir.create("data")
}
if (!file.exists("data/submission")){
  dir.create("data/submission")
}

#R-Packages
IscaretInstalled <- require("caret")
if(!IscaretInstalled){
    install.packages("caret")
    library("caret")
    }

IsrandomForestInstalled <- require("randomForest")
if(!IsrandomForestInstalled){
    install.packages("randomForest")
    library("randomForest")
    }

IsRpartInstalled <- require("rpart")
if(!IsRpartInstalled){
    install.packages("rpart")
    library("rpart")
    }

IsRpartPlotInstalled <- require("rpart.plot")
if(!IsRpartPlotInstalled){
    install.packages("rpart.plot")
    library("rpart.plot")
    }

# Set seed for reproducability
set.seed(9999)
```

## Data processing
The data is downloaded and processed in this section. So that 'NA' values are excluded, some basic adjustments and cleanup will be performed. In the subset, irrelevant columns like 'user name,' raw timestamp part 1,' raw timestamp part 2,' cvtd timestamp,' new window, and num window' (columns 1 to 7) will be eliminated.

The information in the 'pml-training.csv' file is utilised to create training and testing sets.
Based on the trained model, the pml-test.csv data is utilised to predict and answer the 20 questions.

```{r dataprocessing, echo=TRUE, results='hide'}
# Download data
download.file(training.url, training.file)
download.file(test.cases.url,test.cases.file )

# Clean data
training   <-read.csv(training.file, na.strings=c("NA","#DIV/0!", ""))
testing <-read.csv(test.cases.file , na.strings=c("NA", "#DIV/0!", ""))
training<-training[,colSums(is.na(training)) == 0]
testing <-testing[,colSums(is.na(testing)) == 0]

# Subset data
training   <-training[,-c(1:7)]
testing <-testing[,-c(1:7)]
```

## Cross-validation
Cross-validation will be performed in this part by splitting the training data into training (75%) and testing (25%) data.

```{r datasplitting, echo=TRUE, results='hide'}
subSamples <- createDataPartition(y=training$classes, p=0.75, list=FALSE)
subTraining <- training[subSamples,] 
subTesting <- training[-subSamples,]
```

## Expected out-of-sample error
The quantity: 1-accuracy in the cross-validation data will correspond to the expected out-of-sample error. The proportion of correctly identified observations in the subTesting data set is called accuracy. The expected accuracy in the out-of-sample data set is called expected accuracy (i.e. original testing data set). As a result, the expected value of the out-of-sample error will be the quantity: 1-accuracy found from the cross-validation data set, which is the expected number of missclassified observations/total observations in the Test data set.

## Exploratory analysis
The variable `classes` contains 5 levels. The plot of the outcome variable shows the frequency of each levels in the subTraining data.

```{r exploranalysis, echo=TRUE}
plot(subTraining$classes, col="orange", main="Levels of the variable classes", xlab="classes levels", ylab="Frequency")
```

The plot above shows that Level A is the most frequent classes. D appears to be the least frequent one.

## Prediction models
In this section a decision tree and random forest will be applied to the data.

### Decision tree
```{r decisiontree, echo=TRUE}
# Fit model
modFitDT <- rpart(classes ~ ., data=subTraining, method="class")

# Perform prediction
predictDT <- predict(modFitDT, subTesting, type = "class")

# Plot result
rpart.plot(modFitDT, main="Classification Tree", extra=102, under=TRUE, faclen=0)
```

Following confusion matrix shows the errors of the prediction algorithm.

```{r decisiontreecm, echo=TRUE}
confusionMatrix(predictDT, subTesting$classes)
```

### Random forest
```{r randomforest, echo=TRUE}
# Fit model
modFitRF <- randomForest(classes ~ ., data=subTraining, method="class")

# Perform prediction
predictRF <- predict(modFitRF, subTesting, type = "class")
```

Following confusion matrix shows the errors of the prediction algorithm.

```{r randomforestcm, echo=TRUE}
confusionMatrix(predictRF, subTesting$classes)
```

## Conclusion

### Result

The confusion matrices show, that the Random Forest algorithm performens better than decision trees. The accuracy for the Random Forest model was 0.995 (95% CI: (0.993, 0.997)) compared to 0.739 (95% CI: (0.727, 0.752)) for Decision Tree model. The random Forest model is choosen.

### Expected out-of-sample error
The expected out-of-sample error is estimated at 0.005, or 0.5%. The expected out-of-sample error is calculated as 1 - accuracy for predictions made against the cross-validation set. Our Test data set comprises 20 cases. With an accuracy above 99% on our cross-validation data, we can expect that very few, or none, of the test samples will be missclassified.

## Submission
In this section the files for the project submission are generated using the random forest algorithm on the testing data.

```{r submission, echo=TRUE}
# Perform prediction
predictSubmission <- predict(modFitRF, testing, type="class")
predictSubmission

# Write files for submission
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("./data/submission/problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictSubmission)
